# Cooperative binding models for DNA-protein interactions

Edward Wallace, based on Aarti Venkat's R code.

In R markdown format: to compile and run code, open .Rmd file in Rstudio and press "Knit HTML"

## Load R packages
    
```{r}
library("nls2")
library("reshape2")
library("ggplot2")
library("plyr")
theme_set(theme_bw())
```

## Load sum-normalized data and assemble into data frame
    
```{r}
#5.6 uM data ancerr erre61pal, here it is
newconc<- c( 0.0,2.7,5.4,11.7,26.1,38.7,58.5,88.2,197.1,444.6,666.9,1500.3,2250.0,3375.0,5062.5)
# newconc <- newconc*0.9 # why multiply by 0.9?
fbound0 <- c(1,1,1, 0.994,0.989,0.968,0.924,0.634,0.032,0.0139,0,0,0,0,0)
fbound1 <- c(0,0,0,0.006,0.0103,0.0315,0.0752, 0.313,0.042,0.007,0,0,0,0,0)
fbound2 <- c(0, 0, 0, 0, 0, 0, 0, 0.0523,0.925,0.978, 1, 1, 1, 1, 1)

# package into "wide" data frame with one column per variable, one row per concentration
erre61pal_df_wide <- data.frame(conc=newconc,fbound0,fbound1,fbound2)


```

## Functions to reshape and plot data

```{r}
make_fbounddf_long <- function(fbdf) {
    # reshape into "long" dataframe with one row per observation
    # output is almost identical to "combineddata"
    melteddf <- melt(fbdf,id.vars="conc",value.name ="fraction_bound",variable.name="n_prot")
    # awkward conversion of n_prot from name (e.g. fbound0) to n_prot count (e.g. 0)
    melteddf$n_prot <- as.integer(gsub("fbound","",as.character(melteddf$n_prot)))
    return(melteddf)
    }

plot_fbound <- function(fbound_df, title="fbound vs protein concentration") {
    ggplot(data=fbound_df,aes(x=conc,y=fraction_bound,colour=factor(n_prot))) +
        geom_point() +
        scale_x_log10() +
        scale_colour_manual("n bound", values=c("0"="red","1"="blue","2"="green2")) +
        labs(title=title,x="concentration (uM?)",y="fraction bound")
    }
```

## Actually plot the data

```{r,fig.width=9,fig.height=4}
erre61pal_df <- make_fbounddf_long(erre61pal_df_wide)
str(erre61pal_df)
head(erre61pal_df)

plot_fbound(erre61pal_df,"AncERR ERRE61Pal")
```

## Function to fit data to Brenowitz 1991 model

Tested by fitting with made-up parameters.

```{r,fig.width=9,fig.height=4}
combined_fit_natscale <- function(conc,n_prot,ki,k12,conc.min){
    conc.adj <- pmax(conc - conc.min, 0 ) # Why is this needed?
    Z <- (1 + 2*ki*conc.adj + ki^2*k12*conc.adj^2)
    fbound_fit_unnorm <-  
        (2*ki*conc.adj) ^ (n_prot==1) * 
        (ki^2*k12*conc.adj^2) ^(n_prot==2) 
    fbound_fit_unnorm/Z
    }


plot_fbound_fit <- function(fbound_df, title="fbound vs protein concentration",
                            mycoef=c(ki=0.01,k12=1,conc.min=20),
                            interpolate=TRUE,i_factor=10) {
    # plot fraction bound AND predicted fraction bound from coefficients mycoef
    
    # automatically add params to title
    newtitle <- paste(title,
                      "\nki=",round(mycoef[1],3),
                      ", k12=",round(mycoef[2],3),
                      ", conc.min=",round(mycoef[3],1),sep="")
    
    if (!interpolate) {
        # predicted fbound with supplied concentrations only
        fbound_pred_df <- fbound_df
        fbound_pred_df$fraction_bound <- NULL
    } else {
        # smoother predicted fbound with intermediate concentrations
        fbound_pred_df <- ddply(fbound_df,~n_prot, function(df) {
            minconc <- min(fbound_df$conc[fbound_df$conc > 0])
            maxconc <- max(fbound_df$conc[fbound_df$conc > 0])
            # i_factor times as many points as the original data, evenly on logscale
            npoints <- i_factor * nrow(df)
            concout <- c(0,exp(seq(log(minconc),log(maxconc), length.out=npoints)))
            data.frame(conc=concout)
            })
        }
    # calculate predicted fbound
   fbound_pred_df$fraction_bound <- with(fbound_pred_df,combined_fit_natscale(conc,n_prot,
                                           ki=mycoef[1],k12=mycoef[2],conc.min=mycoef[3]))
            
   plot_fbound(fbound_df, newtitle) + 
            geom_line(data=fbound_pred_df,linetype="dashed")
}

plot_fbound_residuals <- function(fbound_df, title="fbound vs protein concentration",
                                  mycoef=c(ki=0.01,k12=1,conc.min=20),
                                  interpolate=TRUE,i_factor=10) {
    # plot fraction bound AND predicted fraction bound from coefficients mycoef
    
    # automatically add params to title
    newtitle <- paste(title,
                      "\nki=",round(mycoef[1],3),
                      ", k12=",round(mycoef[2],3),
                      ", conc.min=",round(mycoef[3],1),sep="")
    # calculate predicted fbound
    fbound_df$predicted_fraction_bound <- with(fbound_df,combined_fit_natscale(conc,n_prot,
                                     ki=mycoef[1],k12=mycoef[2],conc.min=mycoef[3]))
    
    ggplot(data=fbound_df,aes(x=conc,y=fraction_bound-predicted_fraction_bound,colour=factor(n_prot))) +
        geom_point() +
        scale_x_log10() +
        scale_colour_manual("n bound", values=c("0"="red","1"="blue","2"="green2")) +
        labs(title=newtitle,x="concentration (uM?)",y="Dfbound, obs. - pred.")
    }

# test function and fit with arbitrary parameters   
plot_fbound_fit(erre61pal_df,"AncERR ERRE61Pal") 
plot_fbound_residuals(erre61pal_df,"AncERR ERRE61Pal") 
```

## Fit the data by nls and plot the fit

```{r,fig.width=9,fig.height=4}
# initial values

combined_model <- fraction_bound ~ combined_fit_natscale(conc,n_prot,ki,k12,conc.min)

# brute force fit (what is this for?)
start_params <- data.frame(ki=c(0.001,0.1),k12=c(1,20),conc.min=c(15,150))
fit_brute <- nls2(combined_model,data=erre61pal_df,start=start_params, algorithm="brute-force")

# iterative non-linear least squares fit
start_param <- data.frame(ki=0.01,k12=1,conc.min=20)
fit_combined_Gmin_nls2 <- nls2(combined_model,start=start_param,data=erre61pal_df,
                               control=nls.control(warnOnly=TRUE, maxiter=1000))

plot_fbound_fit(erre61pal_df,"AncERR ERRE61Pal, nls fit",mycoef=coef(fit_combined_Gmin_nls2))
plot_fbound_residuals(erre61pal_df,"AncERR ERRE61Pal, nls fit",mycoef=coef(fit_combined_Gmin_nls2))

```

## Do next
* Simulate data with noise
* Try not-sum-normalized data


